### 2007-01-12 Documentation/Paper2/paper2.tex
*** empty log message ***

\newcommand{\GLOBES}{{\sf GLoBES}}
\newcommand{\AEDL}{{\sf AEDL}}
\newcommand{\EDM}{{\sf EDM}}

@@

{\begin{center}
{\Large\bf New features in the simulation of neutrino oscillation experiments with GLoBES 3.0}
\end{center}}
{\begin{center}
{\large\bf (General Long Baseline Experiment Simulator)}
\end{center}}
\renewcommand{\thefootnote}{\alph{footnote}}

@@

\begin{abstract}
We present Version 3.0 of the GLoBES (``General Long Baseline Experiment Simulator'') software, which is
a simulation tool for short- and long-baseline neutrino oscillation experiments. As new features, \GLOBES\ 3.0 allows the simulation of experiments with multiple detectors by the concept of user-defined systematics.
In addition, the combination with external information, such as from different experiment classes, is 
simplified by the concept of user-defined priors. As far as the probability calculation is concerned, \GLOBES\ 
now provides an interface for the inclusion of non-standard physics without re-compliation of the software. 
The set of experiment descriptions coming with \GLOBES\ has been updated. For example, built-in fluxes are now
provided for the simulation of beta beams.
\end{abstract}

@@

The \GLOBES\ software has, in the past, been used for many studies, some of them
are referred to in the paper. However, recent developments have indicated that extensions
and improvements are necessary. In this work, we present the most important new
features and changes in \GLOBES\ 3.0. For experimentalists, \GLOBES\ now allows
for the implementation of arbitrary systematics, which can also be used for
the simulation of multi-detector experiments. In addition, several extensions
have been included in \AEDL\ (``Abstract Experiment Description Language''), such as
built-in beta beam fluxes and the possibility to use lists as variables.
For phenomenologists, user-defined
priors provide a flexible interface to include external information, such as from
different experiments. In addition, \GLOBES\ now can be used for the simulation
of non-standard physics without re-compiling the \GLOBES\ software.

@@

\GLOBES\ allows to simulate experiments with stationary neutrino point 
sources, where each experiment is assumed to have only one neutrino source.
Such experiments are neutrino beam experiments and reactor experiments. 
Geometrical effects of a continuous source distribution, such as in the sun or the 
atmosphere, can not be described. In addition, sources with a physically 
significant time dependencies can not be studied, such as  supernov\ae. 
However, in \GLOBES\ 3.0 and higher, new flexibility is introduced by the
concept of user-defined systematics. This new feature allows the cross-definition
of systematical errors over different experiments. In principle, 
this mechanism can be used for the combination of several discrete sources
and one detector, or several detectors and one source, or several sources and
several detectors. In addition, already implemented
concepts of \GLOBES\ have been used for indirect simulations of geometrical
effects. For example, the mapping of the detector location on the neutrino
energy has been simulated in \Ref~\cite{Rolinec:2006xr} by the use of
variable bin widths.

On the experiment definition side, either built-in neutrino fluxes
(\eg, neutrino factory, beta beam) or arbitrary, user-defined fluxes can be used. 
Similarly,
arbitrary cross sections, energy dependent efficiencies,
energy resolution functions as well as the considered oscillation channels, 
backgrounds, and many other properties can be specified. 
For the systematics, energy
normalization and calibration errors can be simulated, or the
systematics can be completely user-defined. Note that
energy ranges and windows and bin widths can be
(almost) arbitrarily chosen, including bins of different
widths. Together with \GLOBES\ comes a number of
pre-defined experiments in order to demonstrate the capabilities
of \GLOBES\ and to provide prototypes for new experiments.
In addition, they can be used to test new physics ideas with
complete experiment simulations.
% Examples for these prototypes are the \MINOS , \ICARUS , and
% \OPERA\ simulations from \Ref~\cite{Huber:2004ug}, the
% \JHFSK\ and \NUMI\ superbeam simulations from \Refs~\cite{Huber:2002mx,
% Huber:2002rs}, the \JHFHK\ superbeam upgrade simulation from
% \Ref~\cite{Huber:2002mx}, the neutrino factory simulations from
% \Ref~\cite{Huber:2002mx}, and the reactor experiment simulations from
% \Ref~\cite{Huber:2003pm}. Other projects using earlier versions
% of the \GLOBES\ software include \Refs~\cite{Huber:2003ak,Ohlsson:2003ip,Winter:2003ye,Antusch:2004yx}.

@@

\section{User-defined systematics and changes in the systematics implementation}

How defined, what good for (physics examples: T2KK, Reactor-paper,).
Compare to standard systematics (nuisance parameters)

\section{User-defined priors}

@@

\section{Simulation of non-standard physics with GLoBES}

@@

In \GLOBES\ 3.0, we provide an alternative method, which does the full minimization
in one go and is therefore several times faster in most cases. In this algorithm,
each Powell iteration over oscillation parameters is followed by one iteration
over systematics parameters. Since there may be some pathological cases where this
interleaved minimization strategy behaves differently than the old method,
it is by default not used in \GLOBES\ 3.0. This ensures maximal compatibility
with old application programs, but it requires users who wish to benefit from the
new algorithm to explicitly select it by an API function call.

>>>>

\newcommand{\GLOBES}{{\sf GLoBES}}
\newcommand{\GLOBESN}{{\sf GLoBES~3.0}}
\newcommand{\AEDL}{{\sf AEDL}}
%\newcommand{\EDM}{{\sf EDM}}

@@

{\begin{center}
{\Large\bf New features in the simulation of neutrino oscillation experiments with GLoBES 3.0}

(or: ``GLoBES 3.0 release: From user-defined systematics to the simulation of non-standard physics'')
\end{center}}
{\begin{center}
{\large\bf (General Long Baseline Experiment Simulator)}
\end{center}}
\renewcommand{\thefootnote}{\alph{footnote}}

@@

\begin{abstract}
We present Version 3.0 of the GLoBES (``General Long Baseline Experiment Simulator'') software, which is
a simulation tool for short- and long-baseline neutrino oscillation experiments. As new features, \GLOBESN\ allows the simulation of experiments with multiple discrete sources and detectors by the 
concept of user-defined systematics.
In addition, the combination with external information, such as from different experiment classes, is 
simplified by the concept of user-defined priors. As far as the probability calculation is concerned, \GLOBES\ 
now provides an interface for the inclusion of non-standard physics without re-compliation of the software. 
The set of experiment descriptions coming with \GLOBES\ has been updated. For example, built-in fluxes are now
provided for the simulation of beta beams.
\end{abstract}

@@

The \GLOBES\ software has, in the past, been used for many studies, some of them
are referred to in the paper. However, recent developments have indicated that extensions
and improvements are necessary. In this work, we present the most important new
features and changes in \GLOBESN . For experimentalists, \GLOBES\ now allows
for the implementation of arbitrary systematics, which can also be used for
the simulation of multi-detector experiments. In addition, several extensions
have been included in \AEDL\ (``Abstract Experiment Description Language''), such as
built-in beta beam fluxes and the possibility to use lists as variables.
For phenomenologists, user-defined
priors provide a flexible interface to include external information, such as from
different experiments. In addition, \GLOBES\ now can be used for the simulation
of non-standard physics without re-compiling the \GLOBES\ software.

@@

\GLOBES\ allows to simulate experiments with stationary neutrino point 
sources, where each experiment is assumed to have only one neutrino source.
Such experiments are neutrino beam experiments and reactor experiments. 
Geometrical effects of a continuous source distribution, such as in the sun or the 
atmosphere, can not be described. In addition, sources with a physically 
significant time dependencies can not be studied, such as  supernov\ae. 
However, in \GLOBESN\ and higher, new flexibility is introduced by the
concept of user-defined systematics. This new feature allows the cross-definition
of systematical errors over different experiments. In principle, 
this mechanism can be used for the combination of several discrete sources
and one detector, or several detectors and one source, or several sources and
several detectors. In addition, already implemented
concepts of \GLOBES\ have been used for indirect simulations of geometrical
effects. For example, the mapping of the detector location on the neutrino
energy has been simulated in \Ref~\cite{Rolinec:2006xr} by the use of
variable bin widths.

On the experiment definition side, either built-in neutrino fluxes
(\eg, neutrino factory, beta beam) or arbitrary, user-defined fluxes can be used. 
Similarly,
arbitrary cross sections, energy dependent efficiencies,
energy resolution functions as well as the considered oscillation channels, 
backgrounds, and many other properties can be specified. 
For the systematics, energy
normalization and calibration errors can be simulated, or the
systematics can be completely user-defined. Note that
energy ranges and windows and bin widths can be
(almost) arbitrarily chosen, including bins of different
widths. Together with \GLOBES\ comes a number of
pre-defined experiments in order to demonstrate the capabilities
of \GLOBES\ and to provide prototypes for new experiments.
In addition, they can be used to test new physics ideas with
complete experiment simulations.

@@

\section{Systematics implementation and user-defined systematics}

\GLOBES\ supports four types of systematical errors as a standard: Signal and background
normalization errors, as well as signal and background tilt or energy calibration errors.
The tilt (T) is impemented as a linear distortion of the spectrum around the center,
and the energy calibration (C) as a distortion (stretching) of the reconstructed energy scale.
For the implementation of these systematical errors, the  ``pull method'' is used~\cite{Fogli:2002pt}, which introduces  so-called nuisance parameters $\zeta_i$. 
For example, for the signal and background normalization errors, the rates in
each bin $i$ are multiplied with
\begin{equation}
s_i(\zeta_1) := (1+\zeta_1)\cdot s_i \, , \quad b_i(\zeta_2) : =(1+\zeta_2)\cdot b_i \, ,
\end{equation} 
\ie, the rates are scaled by these parameters. The systematics $\chi^2$ is then minimized over
these parameters:
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( \,
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} \frac{\zeta_j^2}{\sigma_{\zeta_j}^2}
\, \right)\,.
\label{equ:chipull}
\end{equation}
Here $\chi^2(\boldsymbol{\lambda}, \zeta_1, \ldots, \zeta_k)$ is the usual Possonian $\chi^2$
depending on the neutrino oscillation parameters $\boldsymbol{\lambda}$ and the nuisance
parameters $\zeta_i$ (which, for instance, scale the signal or background rates).
In addition, Gaussian penalties $\zeta_j^2/\sigma_{\zeta_j}^2$ are added, where 
$\sigma_{\zeta_j}$ corresponds to the actual systematical error.
Core part of \equ{chipull} is the definition of $\chi^2$. For example, for a background-free measurement and
a signal normalization error only, it reads
\begin{equation}
  \chi^2 (\boldsymbol{\lambda},a) =  \sum_{i=1}^{\textrm{\# of bins}}2
     \left( O_{i} - (1 + a) T_{i} +
       (1 + a) T_{i} \log \frac{(1 + a) T_{i}}{O_{i}}  \right) + \frac{a^2}{\sigma_a^2} 
\label{equ:chistandard}
\end{equation}
(JK/PH: CHECK THIS EQUATION IF THIS IS THE ACTUAL IMPLEMENTATION. HERE THE SYSTEMATICS IS\\
IMPLEMENTED IN THE SIMULATED RATES. IN ADDITION, CHECK IF THEORETICAL AND EXP. RATES
ARE CORRECTLY GIVEN. [WW])

Let us assume that we have two detectors, such as for a reactor experiment with a near and
far detector. Using standard systematics in \GLOBES , one has
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}) = \min_{a } \,\, \left( \,
\chi_1^2(\boldsymbol{\lambda},a)+ \frac{a^2}{\sigma_{a}^2} \, \right)
+\min_{ b } \,\, \left( \,
\chi_2^2(\boldsymbol{\lambda},b)+ \frac{b^2}{\sigma_{b}^2}
\, \right)\,.
\label{equ:chipullex}
\end{equation}
In this case, the normalization between the two detectors will be uncorrelated.

Using \GLOBESN\ and higher, one can modify the standard $\chi^2$ function in
\GLOBES\ which corresponds to \equ{chistandard}. In addition, the systematics
concept of ``error dimensions'' is replaced by a concept directly related to
this $\chi^2$ function. As an example, consider a reactor experiment with near ($N$)
and far ($F$) detectors which are described by a reactor flux uncertainty $\sigma_R$
and two fiducial volume errors of the near $\sigma_N$ and far $\sigma_F$ detectors.
Then \equ{chistandard} will be replaced by
\begin{align}
  \chi^2(\boldsymbol{\lambda},a_N,a_F,a_R) = & \sum_{i=1}^{\textrm{\# of bins}} \sum_{d = N,F} 2
     \left( O_{d,i} - (1 + a_R + a_d) T_{d,i} + \right. \nonumber \\
& \left. + (1 + a_R + a_d) T_{d,i} \log \frac{(1 + a_R + a_d) T_{d,i}}{O_{d,i}}  \right) 
          + \frac{a_R^2}{\sigma_R^2} + \frac{a_N^2}{\sigma_N^2} + \frac{a_F^2}{\sigma_F^2}
\label{equ:chi2reactor}
\end{align}
HOW CAN ONE MAKE THE LEFT BRACKET LARGER? [WW]
This function can be defined with user-defined systenatics. Similarly, calibration errors, shape
errors, uncorrelated bin-to-bin errors, \etc , can be introduced. Furthermore, more complicated setups using multiple discrete sources and detectors can be simulated. For details, we refer to the manual~\cite{Manual}.
In the past, user-defined systematics has (as experimental feature) been used in \Ref~\cite{Huber:2006vr} to
simulate Double Chooz and in \Ref~\cite{Barger:2006kp} to simulate T2KK. ANYTHING ELSE? [WW]

In summary, \GLOBES\ currently supports the following systematics $\chi^2$ functions:
\begin{description}
\item[{\tt chiSpectrumTilt}]
 Spectral information, signal and background normalization errors and tilts 
\item[{\tt chiNoSysSpectrum}]
 No systematical errors, but spectral information
\item[{\tt chiTotalRatesTilt}]
 Total rates only, signal and background normalization errors and tilts 
\item[{\tt chiSpectrumOnly}]
 Spectrum only, no other systematics 
\item[{\tt chiNoSysTotalRates}]
 Total rates only, no systematics
\item[{\tt chiSpectrumCalib}]
 Spectral information, signal and background normalization errors and energy calibration errors 
\item[{\tt chiZero}]
 Passive systematics (returns $\chi^2=0$, only provides the event rates for the access by other
rules or experiments)
\item[{\tt Any other name}]
 User-defined systematics
\end{description}

\section{Adding external information with user-defined priors}

@@

\section{Simulation of non-standard physics}

@@

In \GLOBESN , we provide an alternative method, which does the full minimization
in one go and is therefore several times faster in most cases. In this algorithm,
each Powell iteration over oscillation parameters is followed by one iteration
over systematics parameters. Since there may be some pathological cases where this
interleaved minimization strategy behaves differently than the old method,
it is by default not used in \GLOBESN . This ensures maximal compatibility
with old application programs, but it requires users who wish to benefit from the
new algorithm to explicitly select it by an API function call.
